An Efficient Minibatch Acceptance Test for Metropolis-Hastings

### 2017-01-09 arXiv.tex
added arxiv.tex back

% if you need to pass options to natbib, use, e.g.: \PassOptionsToPackage{numbers, compress}{natbib}
% before loading nips_2016
% to avoid loading the natbib package, add option nonatbib: \usepackage[nonatbib]{nips_2016}

\PassOptionsToPackage{numbers, compress}{natbib}
%\usepackage{nips_2016}
% to compile a camera-ready version, add the [final] option, e.g.:
\usepackage[final]{nips_2016}

\usepackage[utf8]{inputenc} % allow utf-8 input
\usepackage[T1]{fontenc}    % use 8-bit T1 fonts
\usepackage{hyperref}       % hyperlinks
\usepackage{url}            % simple URL typesetting
\usepackage{booktabs}       % professional-quality tables
\usepackage{amsfonts}       % blackboard math symbols
\usepackage{amssymb}       % blackboard math symbols
\usepackage{nicefrac}       % compact symbols for 1/2, etc.
\usepackage{microtype}      % microtypography
\usepackage{graphicx}
\usepackage{caption}
\usepackage{subcaption}

\usepackage{amsmath,amsthm,color,graphicx,verbatim,listings,enumitem}
\usepackage[]{algorithm2e}
\graphicspath{{figures/}}

\lstset{
numbers=left, 
numberstyle=\small, 
numbersep=8pt, 
frame = single, 
language=matlab, 
framexleftmargin=20pt}

\newcommand{\simiid}{\overset{\textrm{i.i.d.}}{\sim}}
\DeclareMathOperator*{\argmin}{arg\,min}
\newtheorem{lemma}{Lemma}
\newtheorem{theorem}{Theorem}
\newtheorem{corollary}{Corollary}

@@

% The \author macro works with any number of authors. There are two commands used to separate the
% names and addresses of multiple authors: \And and \AND.
% Using \And between authors leaves it to LaTeX to determine where to break the lines. Using \AND
% forces a line break at that point. So, if LaTeX puts 3 of 4 authors names on the first line, and
% the last on the second line, try using \AND instead of \And before the third author name.

%%  David S.~Hippocampus\thanks{Use footnote for providing further
%%    information about author (webpage, alternative
%%    address)---\emph{not} for acknowledging funding agencies.} \\
%%  Department of Computer Science\\
%%  Cranberry-Lemon University\\
%%  Pittsburgh, PA 15213 \\
%%  \texttt{hippo@cs.cranberry-lemon.edu} \\

\author{
  Haoyu Chen\textsuperscript{*}, Daniel Seita\textsuperscript{*},
  Xinlei Pan\textsuperscript{*}, John Canny\textsuperscript{*}$^{\dag}$ \\
  * University of California, Berkeley, CA \\
$\dag$ Google Research, Mountain View, CA\\
  \texttt{\{haoyuchen,seita,xinleipan,canny\}@berkeley.edu}
 }

\begin{document}
\maketitle

\begin{abstract}
Markov chain Monte Carlo (MCMC) methods have many applications in
machine learning. We are particularly interested in their application
to modeling very large datasets, where it is impractical to perform
Metropolis-Hastings tests on the full data. Previous work on reducing
the cost of Metropolis-Hastings tests yield variable data consumed per
sample, with only constant factor reductions versus using the full
dataset for each sample.  Here we present a method that can be tuned
to provide arbitrarily small batch sizes, by adjusting either proposal
step size or temperature. Our approach uses the natural noise present
in minibatch likelihood estimates to furnish the randomness in a
Metropolis-Hastings test. Our test uses the noise-tolerant Barker
acceptance test with a novel additive correction variable.  The
resulting test can be combined with minibatch proposals to yield
updates with the same complexity as a simple SGD update. In this paper
we derive the test, analyze its performance, discuss its
implementation, and present several experiments. We demonstrate 
several order-of-magnitude speedup over previous work, and show for
the first time that small expected minibatch sizes are possible. 
\end{abstract}

\section{Introduction}\label{sec:introduction}

Markov chain Monte Carlo (MCMC) sampling is a powerful method for
computation on intractable distributions. We are interested primarily
in large dataset applications, where the goal is to sample a posterior
distribution $p(\theta \mid x_1, \ldots, x_N)$ of parameter $\theta$,
and where the number of data instances $N$ is large.  The
Metropolis-Hastings method (M-H) generates sample candidates from a
proposal distribution $q$ which is in general different from the
target distribution $p$, and decides whether to accept or reject them
based on an acceptance test. The acceptance test is usually a
Metropolis test~\cite{Metropolis1953,hastings70}. Conventional
Metropolis-Hastings requires all $N$ data instances to generate one
posterior sample.

Many state-of-the-art machine learning methods, and deep learning in
particular, are based on minibatch updates (such as SGD) to a model.
Minibatch updates produce many improvements to the model for each pass
over the dataset, and have high sample efficiency. They also map very
well onto hardware such as GPUs. In contrast, M-H requires
calculations over the full dataset to produce a new sample.  Recent
results from~\cite{cutting_mh_2014,icml2014c1_bardenet14} perform
approximate (bounded error) acceptance tests using subsets
(minibatches) of the full dataset. The tests depend on minibatch
statistics, and on the value of an additional uniform random variable
$u$. The amount of data consumed for each test varies significantly
from one minibatch to the next, and depends on the current sample, the
proposed sample, and on the random variable $u$. By
contrast,~\cite{conf/uai/MaclaurinA14,TallData15} perform exact
tests but require a lower bound on parameter distribution across its
domain.  The amount of data reduction depends on the accuracy of this
bound, and such bounds are only available for relatively simple
distributions. By contrast our approach is ``black box'' and requires
only statistics gathered from the minibatches as they are used.

Here we derive a new test which incorporates the variability in minibatch
statistics as {\em a natural part of the test}. Because of this, the amount of
data required for each test is fixed with high probability and in expected
value.
%% {\color{blue} Daniel: we really need to show the amount of data is fixed
%% with ``high probability'' (the results suggest ``OK probability'').} 
%% Analysis show data amount is gaussian and short-tailed and therefore bounded with high p
%% experiments back this up. 
We use a Barker test function~\cite{Barker65} rather than a Metropolis
test, which makes our test naturally error tolerant. The idea of using
a noise-tolerant test using Barker's test function was suggested but
not explored empirically in~\cite{TallData15} section 6.3. But the
asymptotic test statistic CDF and the Barker function are different,
which leads to fixed errors for the approach
in~\cite{TallData15}. Here we show that the difference between the
distributions can be corrected with an additive random variable. This
leads to a test which is fast, and whose error can be made arbitrarily
small.

Our test is applicable when the variance (over data samples) of the log
acceptance probability is small enough (less than 1). It's not clear at first why
this quantity should be bounded, but we will show that it is ``natural'' for
well-specified models running Metropolis-Hastings sampling with optimal
proposals~\cite{OptimalScaling01} on a full dataset. When we reduce the amount
of data for the test, the variance goes up. We have to reduce variance in one
of several ways. Either:

\begin{itemize}

\item Increase the temperature of the target distribution. Log likelihoods
  scale as $1/T$, and so the variance of the likelihood ratio will
  vary as $1/T^2$. Our model is no longer well-specified (we are doing inference
  at a temperature different from that assumed during data generation), but
  higher temperature can be advantageous for parameter exploration.

\item For continuous probability distributions, reduce the proposal
  step size and variance (for stochastic proposals) compared to an optimal
  proposal. The variance of the log acceptance probability scales as the
  square of proposal step size.

\item Increase the minibatch size when needed for certain
  minibatches. Log acceptance variance scales as $1/k$ vs the
  minibatch size $k$. Our test is adaptive like earlier works, but
  unlike them, the distribution of minibatch size is Gaussian, not
  long-tailed.  Increased minibatch size also reduces the error rate
  for the test.

\end{itemize}

It's worth discussing at this point the typical goals of
M-H sampling on very large datasets.  By the Bernstein-von Mises
Theorem, the posterior distribution of the parameter $\theta$ for a
Bayesian inference task is asymptotically normal, and has variance
that scales inversely with the number of data samples $N$. This mode
is extremely sharp for large datasets, which may contain millions or
billions of samples. Simply sampling from this distribution is one
application, but an efficient proposal distribution~\cite{OptimalScaling01} has similar variance to the target
distribution and will diffuse to it extremely slowly from an
initialization value which is (likely to be) many standard deviations
away. If there are any other strong modes, it is very likely for the
sampler to find one of them and become trapped in it when run at the
normal distribution temperature ($T=1$). A common solution is to anneal
the sampler, running first at high temperature (scaling log
likelihoods by $1/T$) which flattens the likelihood landscape.  This in
turn reduces the variance of the log acceptance probability and allows
our acceptance test to be applied.

A second question concerns step size. Once we have fixed temperature,
our variance constraint implies that we have to trade-off proposal
step size $s$ and batch size $b$ ($b \propto p^2$), i.e. we can make
many small steps, or one large step, with a given batch of data. One
of the primary drivers of this work is our belief in the value of
small steps. For applications to neural networks or other models where
the posterior is multimodal, posterior inference is arguably a search
process. Covering the search space densely with small steps is much
more valuable than few sparse steps toward the nearest optimum. In
this mode, Metropolis-Hastings can be used in similar fashion to
Stochastic Gradient Descent. The goal in SGD is to make gradual
progress to a posterior mode with each step, taking small steps so
that the cumulative displacement has progressively lower variance. A
substantial part of the computational work of MCMC on massive datasets
will be similarly in reaching a stationary distribution, which really
means finding a deep posterior mode. Taking noisy small steps will
nevertheless make steady progress to a posterior mode since their bias
is in that direction. We demonstrate this behavior in our logistic regression
experiments.

The contributions of this paper are as follows:

\begin{itemize}
\item We develop a new, more efficient (in samples per test) minibatch acceptance test with quantifiable error bounds. The test
  uses a novel additive correction variable to implement a Barker test based on minibatch mean and variance. 
\item We analyze the test for accuracy and speed.
\item We compared performance of our new test and prior approaches on several datasets. We demonstrate the test is several orders
  of magnitude more efficient than prior work measured as data consumed per test, and that it does not suffer
  from long-tailed minibatch sizes (up to the dataset size). 
\end{itemize}

@@

In the Metropolis-Hastings method~\cite{gilks1996markov,brooks2011handbook}, a
difficult-to-compute probability distribution $p(\theta)$ is sampled using a
Markov chain $\theta_1,\ldots,\theta_n$. The sample $\theta_{t+1}$ at time $t+1$
is generated using a candidate $\theta'$ from a (simpler) proposal distribution
$q(\theta'\mid \theta_t)$, filtered by an acceptance test. The acceptance test
is usually a Metropolis test. The Metropolis test has acceptance probability:
\begin{equation}\label{eq:traditional}
    \alpha(\theta_t,\theta') = \frac{p(\theta')q(\theta_t \mid \theta')}{p(\theta_t)q(\theta' \mid \theta_t)} \wedge 1
\end{equation}
where $a \wedge b$ denotes $\min(a,b)$.  With probability
$\alpha(\theta_t,\theta')$, we accept $\theta'$ and set $\theta_{t+1} =
\theta'$, otherwise set $\theta_{t+1}=\theta_t$.  The test is often implemented
with an auxiliary random variable $u \sim \mathcal{U}(0,1)$ with a comparison
$u<\alpha(\theta_t,\theta')$; here, $\mathcal{U}(a,b)$ denotes the uniform
distribution on the interval $[a,b]$.  For notational simplicity, we drop the
subscript $t$ for the current sample $\theta_t$ and denote it as $\theta$.

The acceptance test guarantees detailed balance, which means
$p(\theta)p(\theta'\mid\theta) = p(\theta')p(\theta \mid\theta')$,
where $p(\theta'\mid\theta)$ is the probability of a transition from state
$\theta$ to $\theta'$. Here $p(\theta'\mid\theta) =
q(\theta'\mid\theta)\alpha(\theta,\theta')$ so the detailed balance equation
becomes
\begin{equation}\label{eq:detailed_balance2}
    p(\theta)q(\theta'\mid\theta)\alpha(\theta,\theta') = p(\theta')q(\theta\mid\theta')\alpha(\theta',\theta)
\end{equation}
This condition, together with ergodicity, guarantees that the Markov chain has a
unique stationary distribution $\pi(\theta) = p(\theta)$.

For Bayesian inference, we would like to sample from a parameter distribution
for $\theta$ based on some observed data, so the target distribution is
$p(\theta \mid x_1, \ldots, x_N)$. The acceptance probability is now:
\begin{equation}\label{eq:acceptance_probability}
    \alpha(\theta,\theta') = 
    \frac{p_0(\theta')\prod_{i=1}^N p(x_i \mid \theta')q(\theta \mid
    \theta')}{p_0(\theta)\prod_{i=1}^N p(x_i \mid \theta)q(\theta' \mid\theta)}
    \wedge 1
\end{equation}
where $p_0(\theta)$ is a prior, and $p(x_i \mid \theta)$ are the probabilities
of the observations. Computing samples this way requires the use of all $N$
training data points, but this is very expensive for large datasets. To address
this challenge,~\cite{cutting_mh_2014,icml2014c1_bardenet14} perform approximate
Metropolis-Hasting tests using sequential hypothesis testing. During each
iteration, they start with a small minibatch of data and test the hypothesis
that the sample $\theta'$ should be accepted based on an approximate version of
the test $u < \alpha(\theta,\theta')$. If the approximate test cannot make a
decision with sufficient confidence, then the minibatch size is increased and
the test repeats. This process continues until a decision. The bounds depend on
either a asymptotic Central Limit Theorem~\cite{cutting_mh_2014} or a
concentration bound~\cite{icml2014c1_bardenet14}. The latter requires direct
bounds on the log likelihood ratio, which for general distributions requires
knowing $p(x_i \mid \theta)$ and $p(x_i \mid \theta')$ for all $N$ samples. In
addition, both methods suffer the drawback of resolving small log likelihood
ratio differences between the minibatch and full batch versions.  In the worst
case, all $N$ data points may be needed, and we soon show that this worst case
can occur about $\Omega(N)$ times during the performance of $N$ tests.

Following~\cite{icml2014c1_bardenet14}, we write the test $u <
\alpha(\theta,\theta')$ in the equivalent form $\Lambda(\theta,\theta') >
\psi(u,\theta,\theta')$, where\footnote{Our definitions differ from those in
\cite{icml2014c1_bardenet14} by a factor of $N$ to simplify our analysis later.}
\begin{equation}
\Lambda(\theta,\theta') = \sum_{i=1}^N \log\left(\frac{p(x_i|\theta')}{p(x_i|\theta)}\right)  
\hspace{0.2in} {\rm and} \hspace{0.2in}
\psi(u,\theta,\theta') = \log\left(u\frac{q(\theta'|\theta)p_0(\theta)}{q(\theta|\theta')p_0(\theta')}\right)
\end{equation}
To reduce computational effort, an unbiased estimate of $\Lambda(\theta,\theta')$
based on a minibatch can be used:
\begin{equation}
\Lambda^*(\theta,\theta') = \frac{N}{b}\sum_{i=1}^b \log\left(\frac{p(x_i|\theta')}{p(x_i|\theta)}\right)  
\end{equation}
Finally, it will be convenient for our analysis to define the individual
components that contribute to the sums above:
\begin{equation}\label{eq:individual_terms}
\Lambda_i(\theta,\theta') = N \log\left(\frac{p(x_i|\theta')}{p(x_i|\theta)}\right)  
\end{equation}
Thus, $\Lambda(\theta,\theta')$ is the mean of $\Lambda_i(\theta,\theta')$ over
the entire dataset, and $\Lambda^*(\theta,\theta')$ is the mean of
$\Lambda_i(\theta,\theta')$ over its minibatch.

Since minibatches contains randomly selected samples $x_i$, the values
$\Lambda_i$ are independent, identically distributed (iid) random
variables\footnote{The analysis assumes sampling with replacement
although implementations on typical large datasets will approximate
this by sampling without replacement.}.
By the Central Limit Theorem, we expect $\Lambda^*(\theta,\theta')$ to
be approximately Gaussian. The acceptance test then becomes a
statistical test of the hypothesis that
$\Lambda(\theta,\theta')>\psi(u,\theta,\theta')$ by establishing that
$\Lambda^*(\theta,\theta')$ is substantially larger than
$\psi(u,\theta,\theta')$.  In~\cite{cutting_mh_2014} an asymptotic
central limit argument was used to derive this gap, while
in~\cite{icml2014c1_bardenet14} a concentration bound was used. In
both cases, the resulting tests were shown to give useful reductions
in number of samples required over using the full dataset, but there
were no worst-case bounds given on the average number of samples
needed per iteration.

We next show that for a simple distribution, the lower bound of average
number of data instances consumed for one iteration of~\cite{cutting_mh_2014}
and~\cite{icml2014c1_bardenet14} is $\Omega(N)$, where $N$ is the number of data
points.

\subsection{A Worst-Case Gaussian Example}\label{ssec:gaussian_example}
Consider a Gaussian model where $x_1,\ldots,x_N \simiid \mathcal{N}(\theta,1)$
with a known variance $\sigma^2=1$, true mean $\theta=0.5$, and a uniform prior
on $\theta$. The log likelihood ratio is
\begin{equation}\label{eq:lemma_ll_ratio}
\Lambda^*(\theta,\theta') =  \frac{N}{b}\sum_{i=1}^b \log\frac{p(x_i|\theta')}{p(x_i|\theta)}=
  N(\theta'-\theta)\left(\frac{1}{b}\sum_{i=1}^b x_i-\theta-\frac{\theta'-\theta}{2}\right)
\end{equation}
which is normally distributed over selection of the Normal samples $x_i$.  Since
the $x_i$ have unit variance, their mean has variance $1/b$, and the variance of
$\Lambda^*(\theta,\theta')$ is $\sigma^2(\Lambda^*) = (\theta'-\theta)^2N^2/b$.
In order to pass a hypothesis test that $\Lambda > \psi$, there needs to be a
large enough gap (several $\sigma(\Lambda^*)$) between
$\Lambda^*(\theta,\theta')$ and $\psi(u,\theta,\theta')$.

>>>>

\usepackage[final]{nips_2016}

% If your paper is accepted, change the options for the package
% aistats2017 as follows:
%
%\usepackage[accepted]{aistats2017}
%
% This option will print headings for the title of your paper and
% headings for the authors names, plus a copyright note at the end of
% the first column of the first page.
\usepackage{amsmath,amsthm,color,graphicx,verbatim,listings,enumitem,float}
\usepackage{amssymb}
\usepackage{natbib}  % for references
\usepackage{amsfonts}
\usepackage{nicefrac}  
\usepackage{booktabs}
\usepackage{microtype}
\usepackage[]{algorithm2e}
\usepackage{caption}
\usepackage{subcaption}
\graphicspath{{figures/}}
\usepackage{setspace}
\lstset{
numbers=left, 
numberstyle=\small, 
numbersep=8pt, 
frame = single, 
language=matlab, 
framexleftmargin=20pt}

\renewcommand{\refname}{}
\newcommand{\simiid}{\overset{\textrm{i.i.d.}}{\sim}}
\DeclareMathOperator*{\argmin}{arg\,min}
\newtheorem{lemma}{Lemma}
\newtheorem{theorem}{Theorem}
\newtheorem{corollary}{Corollary}

% This portion of style definition will remove the index number before the 
% without inducing an indent
\makeatletter
\renewcommand\@biblabel[1]{}
\renewenvironment{thebibliography}[1]
     {\section*{\refname}%
      \@mkboth{\MakeUppercase\refname}{\MakeUppercase\refname}%
      \list{}%
           {\leftmargin0pt
            \@openbib@code
            \usecounter{enumiv}}%
      \sloppy
      \clubpenalty4000
      \@clubpenalty \clubpenalty
      \widowpenalty4000%
      \sfcode`\.\@m}
     {\def\@noitemerr
       {\@latex@warning{Empty `thebibliography' environment}}%
      \endlist}
\makeatother

\begin{document}

% If your paper is accepted and the title of your paper is very long,
% the style will print as headings an error message. Use the following
% command to supply a shorter title of your paper so that it can be
% used as headings.
%
%\runningtitle{I use this title instead because the last one was very long}

% If your paper is accepted and the number of authors is large, the
% style will print as headings an error message. Use the following
% command to supply a shorter version of the authors names so that
% they can be used as headings (for example, use only the surnames)
%
%\runningauthor{Surname 1, Surname 2, Surname 3, ...., Surname n}

@@

\author{ Anonymous Author 1 \And Anonymous Author 2 \And Anonymous Author 3 }

% \address{ Unknown Institution 1 \And Unknown Institution 2 \And Unknown Institution 3 } ]

\maketitle

\begin{abstract}
  %The Abstract paragraph should be indented 0.25 inch (1.5 picas) on
  %both left and right-hand margins. Use 10~point type, with a vertical
  %spacing of 11~points. The {\bf Abstract} heading must be centered,
  %bold, and in point size 12. Two line spaces precede the
  %Abstract. The Abstract must be limited to one paragraph.
  Markov chain Monte Carlo (MCMC) methods have many applications in
  machine learning. We are particularly interested in their application
  to modeling very large datasets, where it is impractical to perform
  Metropolis-Hastings tests on the full data. Previous work on reducing
  the cost of Metropolis-Hastings tests yield variable data consumed per
  sample, with only constant factor reductions versus using the full
  dataset for each sample.  Here we present a method that can be tuned
  to provide arbitrarily small batch sizes, by adjusting either proposal
  step size or temperature. Our approach uses the natural noise present
  in minibatch likelihood estimates to furnish the randomness in a
  Metropolis-Hastings test. Our test uses the noise-tolerant Barker
  acceptance test with a novel additive correction variable.  The
  resulting test can be combined with minibatch proposals to yield
  updates with the same complexity as a simple SGD update. In this paper
  we derive the test, analyze its performance, discuss its
  implementation, and present several experiments. We demonstrate 
  several order-of-magnitude speedup over previous work, and show for
  the first time that small expected minibatch sizes are possible. 
\end{abstract}

\section{Introduction}\label{sec:introduction}
Markov chain Monte Carlo (MCMC) sampling is a powerful method for
computation on intractable distributions. We are interested primarily
in large dataset applications, where the goal is to sample a posterior
distribution $p(\theta \mid x_1, \ldots, x_N)$ of parameter $\theta$,
and where the number of data instances $N$ is large.  The
Metropolis-Hastings method (M-H) generates sample candidates from a
proposal distribution $q$ which is in general different from the
target distribution $p$, and decides whether to accept or reject them
based on an acceptance test. The acceptance test is usually a
Metropolis test~\citep{Metropolis1953, hastings70}. Conventional
Metropolis-Hastings requires all $N$ data instances to generate one
posterior sample.

Many state-of-the-art machine learning methods, and deep learning in
particular, are based on minibatch updates (such as SGD) to a model.
Minibatch updates produce many improvements to the model for each pass
over the dataset, and have high sample efficiency. %They also map very
% well onto hardware such as GPUs. 
In contrast, M-H requires calculations over the full dataset to produce a new sample.  Recent
results from~\citet{cutting_mh_2014} and ~\citet{icml2014c1_bardenet14} perform
approximate (bounded error) acceptance tests using subsets
(minibatches) of the full dataset. The tests depend on minibatch
statistics, and on the value of an additional uniform random variable
$u$. The amount of data consumed for each test varies significantly
from one minibatch to the next, and depends on the current sample, the
proposed sample, and on the random variable $u$. By
contrast,~\citet{conf/uai/MaclaurinA14,TallData15} perform exact
tests but require a lower bound on parameter distribution across its
domain.  The amount of data reduction depends on the accuracy of this
bound, and such bounds are only available for relatively simple
distributions. %By contrast our approach is a ``black box'' and requires
%only statistics gathered from the minibatches as they are used. 
Here we derive a new test which incorporates the variability in minibatch
statistics as {\em a natural part of the test}. Because of this, the amount of
data required for each test is fixed with high probability and in expected
value.

We use a Barker test function~\citep{Barker65} rather than a Metropolis
test, which makes our test naturally error tolerant. The idea of using
a noise-tolerant %test using
Barker's test function was suggested but
not explored empirically in~\citet{TallData15} section 6.3. But the
asymptotic test statistic CDF and the Barker function are different,
which leads to fixed errors for the approach
in~\citet{TallData15}. Here we show that the difference between the
distributions can be corrected with an additive random variable. This
leads to a test which is fast, and whose error can be made arbitrarily
small.

Our test is applicable when the variance (over data samples) of the log
acceptance probability is small enough (less than 1). It's not clear at first why
this quantity should be bounded, but we will show that it is ``natural'' for
well-specified models running Metropolis-Hastings sampling with optimal
proposals~\citep{OptimalScaling01} on a full dataset. When we reduce the amount
of data for the test, the variance goes up. We have to reduce variance in one
of several ways. Either:

\begin{itemize}[noitemsep]
\item Increase the temperature of the target distribution. Log likelihoods
  scale as $1/T$, and so the variance of the likelihood ratio will
  vary as $1/T^2$. Our model is no longer well-specified (we are doing inference
  at a temperature different from that assumed during data generation), but
  higher temperature can be advantageous for parameter exploration.
\item For continuous probability distributions, reduce the proposal
  step size and variance (for stochastic proposals) compared to an optimal
  proposal. The variance of the log acceptance probability scales as the
  square of proposal step size. 
\item Increase the minibatch size when needed for certain
  minibatches. Log acceptance variance scales as $1/k$ vs the
  minibatch size $k$. Our test is adaptive like earlier works, but
  unlike them, the distribution of minibatch size is Gaussian, not
  long-tailed.  Increased minibatch size also reduces the error rate
  for the test.
\end{itemize}

It's worth discussing at this point the typical goals of
M-H sampling on very large datasets.  By the Bernstein-von Mises
Theorem, the posterior distribution of the parameter $\theta$ for a
Bayesian inference task is asymptotically normal, and has variance
that scales inversely with the number of data samples $N$. %This mode
%is extremely sharp for large datasets, which may contain millions or
%billions of samples. 
Simply sampling from this distribution is one
application, but an efficient proposal distribution~\citep{OptimalScaling01} has similar variance to the target
distribution and will diffuse to it extremely slowly from an
initialization value which is (likely to be) many standard deviations
away. If there are any other strong modes, it is very likely for the
sampler to find one of them and become trapped in it when run at the
normal distribution temperature ($T=1$). A common solution is to anneal
the sampler, running first at high temperature (scaling log
likelihoods by $1/T$) which flattens the likelihood landscape.  This in
turn reduces the variance of the log acceptance probability and allows
our acceptance test to be applied.

A second question concerns step size. Once we have fixed temperature,
our variance constraint implies that we have to trade-off proposal
step size $s$ and batch size $b$ ($b \propto p^2$), i.e. we can make
many small steps, or one large step, with a given batch of data. One
of the primary drivers of this work is our belief in the value of
small steps. For applications to neural networks or other models where
the posterior is multimodal, posterior inference is arguably a search
process. Covering the search space densely with small steps is much
more valuable than few sparse steps toward the nearest optimum. In
this mode, Metropolis-Hastings can be used in similar fashion to
Stochastic Gradient Descent. The goal in SGD is to make gradual
progress to a posterior mode with each step, taking small steps so
that the cumulative displacement has progressively lower variance. %A
%substantial part of the computational work of MCMC on massive datasets
%will be similarly in reaching a stationary distribution, which really
%means finding a deep posterior mode. 
Taking noisy small steps will nevertheless make steady progress to 
a posterior mode since their bias is in that direction. We demonstrate 
this behavior in our logistic regression experiments.

The contributions of this paper are as follows:
\begin{itemize}[noitemsep]
\item We develop a new, more efficient (in samples per test) minibatch acceptance test with quantifiable error bounds. The test
  uses a novel additive correction variable to implement a Barker test based on minibatch mean and variance. 
%\item We analyze the test for accuracy and speed.
\item We compared performance of our new test and prior approaches on several datasets. We demonstrate the test is several orders
  of magnitude more efficient than prior work measured as data consumed per test, and that it does not suffer
  from long-tailed minibatch sizes (up to the dataset size). 
\end{itemize}

@@

In the Metropolis-Hastings method~\citep{gilks1996markov,brooks2011handbook}, a
difficult-to-compute probability distribution $p(\theta)$ is sampled using a
Markov chain $\theta_1,\ldots,\theta_n$. The sample $\theta_{t+1}$ at time $t+1$
is generated using a candidate $\theta'$ from a (simpler) proposal distribution
$q(\theta'\mid \theta_t)$, filtered by an acceptance test. The acceptance test
is usually a Metropolis test. The Metropolis test has acceptance probability:
\begin{equation}\label{eq:traditional}
    \alpha(\theta_t,\theta') = \frac{p(\theta')q(\theta_t \mid \theta')}{p(\theta_t)q(\theta' \mid \theta_t)} \wedge 1
\end{equation}
where $a \wedge b$ denotes $\min(a,b)$.  With probability
$\alpha(\theta_t,\theta')$, we accept $\theta'$ and set $\theta_{t+1} =
\theta'$, otherwise set $\theta_{t+1}=\theta_t$.  The test is often implemented
with an auxiliary random variable $u \sim \mathcal{U}(0,1)$ with a comparison
$u<\alpha(\theta_t,\theta')$; here, $\mathcal{U}(a,b)$ denotes the uniform
distribution on the interval $[a,b]$.  For notational simplicity, we drop the
subscript $t$ for the current sample $\theta_t$ and denote it as $\theta$.

The acceptance test guarantees detailed balance, which means
$p(\theta)p(\theta'\mid\theta) = p(\theta')p(\theta \mid\theta')$,
where $p(\theta'\mid\theta)$ is the probability of a transition from state
$\theta$ to $\theta'$. Here $p(\theta'\mid\theta) =
q(\theta'\mid\theta)\alpha(\theta,\theta')$ so the detailed balance equation
becomes 
\begin{equation}\label{eq:detailed_balance2}
    p(\theta)q(\theta'\mid\theta)\alpha(\theta,\theta') = p(\theta')q(\theta\mid\theta')\alpha(\theta',\theta)
\end{equation}
This condition, together with ergodicity, guarantees that the Markov chain has a
unique stationary distribution $\pi(\theta) = p(\theta)$. For Bayesian inference, we would like to sample from a parameter distribution for $\theta$ based on some observed data, so the target distribution is
$p(\theta \mid x_1, \ldots, x_N)$. The acceptance probability is now:
\begin{equation}\label{eq:acceptance_probability}
    \alpha(\theta,\theta') = 
    \frac{p_0(\theta')\prod_{i=1}^N p(x_i \mid \theta')q(\theta \mid
    \theta')}{p_0(\theta)\prod_{i=1}^N p(x_i \mid \theta)q(\theta' \mid\theta)}
    \wedge 1
\end{equation}
where $p_0(\theta)$ is a prior, and $p(x_i \mid \theta)$ are the probabilities
of the observations. Computing samples this way requires the use of all $N$
training data points, but this is very expensive for large datasets. To address
this challenge,~\citet{cutting_mh_2014,icml2014c1_bardenet14} perform approximate
Metropolis-Hasting tests using sequential hypothesis testing. During each
iteration, they start with a small minibatch of data and test whether the sample $\theta'$ 
should be accepted based on an approximate version of
the test $u < \alpha(\theta,\theta')$. If the approximate test cannot make a
decision with sufficient confidence, then the minibatch size is increased and
the test repeats. This process continues until a decision. The bounds depend on
either a asymptotic Central Limit Theorem~\citep{cutting_mh_2014} or a
concentration bound~\citep{icml2014c1_bardenet14}. The latter requires direct
bounds on the log likelihood ratio, which for general distributions requires
knowing $p(x_i \mid \theta)$ and $p(x_i \mid \theta')$ for all $N$ samples. In
addition, both methods suffer the drawback of resolving small log likelihood
ratio differences between the minibatch and full batch versions.  In the worst
case, all $N$ data points may be needed, and we soon show that this worst case
can occur about $\Omega(N)$ times during the performance of $N$ tests.

Following~\citet{icml2014c1_bardenet14}, we write the test $u <
\alpha(\theta,\theta')$ in the equivalent form $\Lambda(\theta,\theta') >
\psi(u,\theta,\theta')$, where\footnote{Our definitions differ from those in
\citet{icml2014c1_bardenet14} by a factor of $N$ to simplify our analysis later.}
\begin{equation}
\begin{split} 
\Lambda(\theta,\theta') & = \sum_{i=1}^N \log\left(\frac{p(x_i|\theta')}{p(x_i|\theta)}\right) \\  
{\rm and} \quad \psi(u,\theta,\theta') &= \log\left(u\frac{q(\theta'|\theta)p_0(\theta)}{q(\theta|\theta')p_0(\theta')}\right) \\
\end{split}
\end{equation}
To reduce computational effort, an unbiased estimate of $\Lambda(\theta,\theta')$
based on a minibatch can be used:
\begin{equation}
\Lambda^*(\theta,\theta') = \frac{N}{b}\sum_{i=1}^b \log\left(\frac{p(x_i|\theta')}{p(x_i|\theta)}\right)  
\end{equation}
Finally, it will be convenient for our analysis to define the individual
components that contribute to the sums above:
\begin{equation}\label{eq:individual_terms}
\Lambda_i(\theta,\theta') = N \log\left(\frac{p(x_i|\theta')}{p(x_i|\theta)}\right)  
\end{equation}
%Thus, $\Lambda(\theta,\theta')$ is the mean of $\Lambda_i(\theta,\theta')$ over
%the entire dataset, and $\Lambda^*(\theta,\theta')$ is the mean of
%$\Lambda_i(\theta,\theta')$ over its minibatch.

Since minibatches contains randomly selected samples $x_i$, the values
$\Lambda_i$ are independent, identically distributed (iid) random
variables\footnote{The analysis assumes sampling with replacement
although implementations on typical large datasets will approximate
this by sampling without replacement.}.
By the Central Limit Theorem, we expect $\Lambda^*(\theta,\theta')$ to
be approximately Gaussian. The acceptance test then becomes a
statistical test of the hypothesis that
$\Lambda(\theta,\theta')>\psi(u,\theta,\theta')$ by establishing that
$\Lambda^*(\theta,\theta')$ is substantially larger than
$\psi(u,\theta,\theta')$.  In~\citet{cutting_mh_2014} an asymptotic
central limit argument was used to derive this gap, while
in~\citet{icml2014c1_bardenet14} a concentration bound was used. In
both cases, the resulting tests were shown to give useful reductions
in number of samples required over using the full dataset, but there
were no worst-case bounds given on the average number of samples
needed per iteration.

We next show that for a simple distribution, the lower bound of average
number of data instances consumed for one iteration of~\citet{cutting_mh_2014}
and~\citet{icml2014c1_bardenet14} is $\Omega(N)$, where $N$ is the number of data
points.

\subsection{A Worst-Case Gaussian Example}\label{ssec:gaussian_example}
Consider a Gaussian model where $x_1,\ldots,x_N \simiid \mathcal{N}(\theta,1)$
with a known variance $\sigma^2=1$, true mean $\theta=0.5$, and a uniform prior
on $\theta$. The log likelihood ratio is
\begin{equation}\label{eq:lemma_ll_ratio}
\resizebox{\linewidth}{!}{% 
$\Lambda^*(\theta,\theta') =  \frac{N}{b}\sum_{i=1}^b \log\frac{p(x_i|\theta')}{p(x_i|\theta)} = N(\theta'-\theta)\left(\frac{1}{b}\sum_{i=1}^b x_i-\theta-\frac{\theta'-\theta}{2}\right)$
}%
\end{equation}
which is normally distributed over selection of the Normal samples $x_i$.  Since
the $x_i$ have unit variance, their mean has variance $1/b$, and the variance of
$\Lambda^*(\theta,\theta')$ is $\sigma^2(\Lambda^*) = (\theta'-\theta)^2N^2/b$.
In order to pass a hypothesis test that $\Lambda > \psi$, there needs to be a
large enough gap (several $\sigma(\Lambda^*)$) between
$\Lambda^*(\theta,\theta')$ and $\psi(u,\theta,\theta')$.
